### 2021-03-24 papers/SYRCoSE/encode_evstr.tex
SYRCoSE: add some text

\subsection{How we encode Event Structures}
% Как мы видели ранее в Step-by-step example наша стуркура событий содержит такие функции как frf fpred. Сами события это элементы некоторого множества E (множества итендификаторов), которое должно удволетворять следущим свойствам:
\begin{definition}[identifier set]
  $\mathcal{E}$ is an identifier set if:
  \begin{itemize}
    \item There is some $ident_0 \in \mathcal{E}$ ~--- first identifier
    \item $\mathcal{E}$ has well-founded partitial order $\leqslant_\mathcal{E}$
    \item There is some function, that takes some finite identifies set $E \subset \mathcal{E}$ and returns a fresh identifier. Formally: $fresh : \mathcal{P}_{fin}(\mathcal{E}) \to \mathcal{E}$, satisfies $$\forall e \in E, \ e <_\mathcal{E} fresh(E)$$
  \end{itemize}
\end{definition}
% Пусть E это множества всех событий программы тогда у нас есть
  $$\ffun{lab} : E \rightharpoonup \lb $$
% где label -- множество всех меток (таких как W(x, 5), R(y, 6), ThreadStart, ThreadEnd etc)
% Мы будем говорить что е это "read"-event c локацией х и значением i если lab(e) = R(x,i), аналогично "write"-event
% В итоге формально мы строим четверку
\begin{definition}[Execution Event Structure]
  $$exec\_event\_structure \ \triangleq \ \langle E, \ \ffun{lab}, \ \fpred, \ \frf \rangle$$
  where
  \begin{itemize}
    \item $E$ ~--- finite subset of some identifiers set $\mathcal{E}$
    \item $\fpred$ ~--- function that returns a predsessor (w.r.t. the program order) of event
    \item $\frf$ ~--- function that takes "read"-event $e$ and returns a "write"-event from witch $e$ reads
    \item $\ffun{lab} : E \to \lb$ ~--- function that returns event's label
  \end{itemize}
  Also we require
  \begin{itemize}
    \item $\forall e \in E, \ \fpred(e) \leqslant_E e$ and $\frf(e) \leqslant_E e$
    \item $\forall e_1 \ e_2 \in E, \ e_2 = \frf(e_1) \ \Leftrightarrow \ e_1$ is a "read"-event and $e_2$ is a "write"-event with the same location and value
  \end{itemize}
\end{definition}

% Далее для простоты будем отождествлять иденфикаторы с их метками, а Execution Event Structure с их множеством итендификаторов

% Теперь наша задача: определить в этих терминах простые структуры событий. Для этого надо определить частичный порядок <= и отношение конфликта #
% Начнем с отношения causality. В нашей статье мы будем считать, что событие e1 влечет событие e2 если 
\begin{itemize}
  \item Programm instruction, that corresponds to event $e_1$ states before the programm instruction, that corresponds to $e_2$
  \item Or $\ e_2$ is the "read"-event and it reads from "write"-event $e_1$
\end{itemize}
% Или формально:
\begin{itemize}
  \item $\fpred(e_1) \ =\ e_2$
  \item Or $\ \frf(e_1) \ =\ e_2$
\end{itemize}
% Любую функцию f можно считать отношением
  $$ a \ f \ b \ \triangleq\ f(a) \ = \ b$$
% Имея ввиду обозначение выше определим
\begin{definition}[causality relation]
  $$\leqslant \ \triangleq \ (\fpred \cup \frf)^* $$
\end{definition}
% Заметим, что в множестве E уже есть тотальный порядок $\leqslant_E$. Нетрудно видеть что полученнле отношение casuality является его предпорядком [смотри Well-foundedness and monads]
\begin{lemma}
  If $\leqslant$ is a casuality order of $E$ then $\leqslant \ \subset \ \leqslant_E$
\end{lemma}
% Теперь давайте определим отношение конфликта. Посмотрев на отношение конфликта в определении СС [ссылка], мы можем установить следущий факт
\begin{lemma}\label{lma_confl}
  If $\ e_1 \leqslant e_2, \ e_3 \leqslant e_4$ and $e_1 \# e_3$ holds then $e_2 \# e_4$
\end{lemma} 
% То есть отношение конфликта должно "протаскиваться" через отношение <=. Наша интуиция будет заключаться в том, что мы определим "минимальное" отношение конфликта, замкнув, котрое по свойству консистентности [ссылка] мы получим полное отношение #. 
% И так откуда берутся конфликты? [тут должен был заход через Step-by-step example]. 
% И так мы видим, что если у событий есть общий предок (относительно отношения fpred) то эти события происходят в двух разных выполнениях программы и, соответвенно находятся в отношении конфликта
\begin{definition}[immediate conflict relation]
  $$ e_1 \, \ffun{icf} \, e_2 \ \triangleq\ \exists e, \, e = \fpred(e_1) = \fpred(e_2)  $$
\end{definition}
% Далее можно определить отношение конфликта как консистентное замыкание icf
\begin{definition}[conflict relation]
  $$ e_1 \, \# \, e_2 \ \triangleq\ \exists e_3\, e_4, e_3 \, \ffun{icf} \, e_4 \ and \ e_3 \leqslant e_1, \, e_4 \leqslant e_2 $$
\end{definition}
% Однако с этим определением возникает одна проблема. Понятно что отношение определенное таким образом получается симметричным и консистентным. Будет ли оно иррефлексивность? Оказывается, что в общем случае нет. Рассмотрим следущую программу:
\input{fig/encode_prog.tex}
% Давайте как в Step-by-step example построим по шагам фрагмент соответвующей этой программе exec_event_structure до чтения из y
\input{fig/encode_part.tex}
% Видно что наша программа в первом потоку разделилась на два независимых выполнения: выполнение в котором мы прочитали из x 1 (обозначим за выболнение (а)) и выполненние в котором прочитали из х 2 (выполнение (b))
%Теперь мы хотим добавить в нашу структуру событий чтение в y. посмотрим что будет если прочитать при выполнении (а) в у 2 (то есть запись из (b))
\input{fig/encode_cf.tex}

@@

% Но отношение конфликта должно быть иррефлексивным. Заметим, однако, что (a) и (b) не могут произойти при одном исполнении программы, а значит и чтения R(y,2) из W(y,2) тут быть не должно, полная структура событий тут должна выглядеть так:
\input{fig/encode_full.tex} \\
% Будем говорить, что сткуктура собырий консистентна если там нет чтений из конфликтующийх с ними записей. Формально
\begin{definition}[consistent Execution Event Structure]
  $$consitent\langle E, \ffun{lab}, \fpred, \frf \rangle \ \triangleq \ \forall e \in E, \ \neg (e \ \# \ \frf e)$$
\end{definition}
% Оказывается этого достатчно чтоб отношение конфликта было иррефлексивным
\begin{lemma}
  Let $E$ be a consistent Execution Event Structure (denote as $cexec\_event\_structure$). Then it's conflict relation $\#$ is irreflexive
\end{lemma}
% Получается что из консистентной exevstr можно получить ПCC

>>>>

\subsection{How we encode Event Structures}
As we saw earlier in Step by Step Example, our event structure contains functions such as $\frf$ and $\fpred$. The events themselves are elements of some set $\mathcal{E}$ (set of identifiers), which must satisfy the following properties
\begin{definition}[identifier set]
  $\mathcal{E}$ is an identifier set if:
  \begin{itemize}
    \item There is some $ident_0 \in \mathcal{E}$ ~--- first identifier
    \item $\mathcal{E}$ has well-founded partitial order $\leqslant_\mathcal{E}$
    \item There is a function, that takes some finite $E \subset \mathcal{E}$ and returns a fresh identifier. Formally: $fresh : \mathcal{P}_{fin}(\mathcal{E}) \to \mathcal{E}$, satisfies $$\forall e \in E, \ e <_\mathcal{E} fresh(E)$$
  \end{itemize}
\end{definition}
We can fix any $\mathcal{E}$ (for example $\mathbb{N}$) and theat programm's events as a finite subset of $\mathcal{E}$

Let $E$ be the set of all program events, then we have
  $$\ffun{lab} : E \to \lb $$
where $\lb$ is the label's set (such that $\wlab{x}{5}, \rlab{y}{6}$ ThreadStart, etc.)

We will say that $e$ is a "read" -event with location x and value i if $\ffun{lab}(e) = \rlab{x}{i}$, similarly to "write" -event

As a result, we formally build a four
\begin{definition}[Execution Event Structure]
  $$exec\_event\_structure \ \triangleq \ \langle E, \ \ffun{lab}, \ \fpred, \ \frf \rangle$$
  where
  \begin{itemize}
    \item $E$ ~--- finite subset of some identifiers set $\mathcal{E}$
    \item $\fpred$ ~--- function that returns a predsessor (w.r.t. the program order) of event
    \item $\frf$ ~--- function that takes "read"-event $e$ and returns a "write"-event from witch $e$ reads
    \item $\ffun{lab} : E \to \lb$ ~--- function that returns event's label
  \end{itemize}
  Also we require
  \begin{itemize}
    \item $\forall e \in E, \ \fpred(e) \leqslant_E e$ and $\frf(e) \leqslant_E e$
    \item $\forall e_1 \ e_2 \in E, \ e_2 = \frf(e_1) \ \Leftrightarrow \ e_1$ is a "read"-event and $e_2$ is a "write"-event with the same location and value
  \end{itemize}
\end{definition}
Further, for simplicity, we will identify identifiers with their labels, and Execution Event Structure with their set of identifiers

Now our goal is to define a prime event structures in these terms. To do this, one need to come up with a causality relation $\leqslant$ and a conflict relation $\#$

Let's start with the causality. In our article, we will assume that the event $e_1$ entails the event $e_2$ if
\begin{itemize}
  \item Programm instruction, that corresponds to event $e_1$ states before the programm instruction, that corresponds to $e_2$
  \item Or $\ e_2$ is the "read"-event and it reads from "write"-event $e_1$
\end{itemize}
Or formally:
\begin{itemize}
  \item $\fpred(e_2) \ =\ e_1$
  \item Or $\ \frf(e_2) \ =\ e_1$
\end{itemize}
Any function $f$ can be treated as a relation
  $$ a \ f \ b \ \triangleq\ f(b) \ = \ a$$
With the notation above, we define

\begin{center}
\begin{definition}[causality relation]
  $$\leqslant \ \triangleq \ (\fpred \cup \frf)^* $$
\end{definition}
\end{center}

Note that the set $E \subset \mathcal{E}$ already contains the total order $ \leqslant_\mathcal{E} $. One can note that the resulting casuality relation is its suborder [TODO:see Well-foundedness and monads]

\begin{center}
\begin{lemma}
  If $\leqslant$ is a casuality order of $E \subset \mathcal{E}$ then $\leqslant \ \subset \ \leqslant_\mathcal{E}$
\end{lemma}
\end{center}

Now let's define a conflict relation. Looking at the conflict relation in \cref{es_def}, we can establish the following fact:
\begin{center}
\begin{lemma}\label{lma_confl}
  If $\ e_1 \leqslant e_2, \ e_3 \leqslant e_4$ and $e_1 \# e_3$ holds then $e_2 \# e_4$
\end{lemma} 
\end{center}
That is, the conflict relation must be "dragged" through the $\leqslant$ relation. Our intuition will be as follows: we, first, define a "minimal" conflict relation, by closing which under the consistency property (\cref{lma_confl}), we finally get a full relation $\#$.

So where do conflicts come from?

From the step-by-step example, we see that if events have a common predsessor (w.r.t. the $\fpred$ relation), then they occur in two different program executions and, accordingly, are conflicting
\begin{definition}[immediate conflict relation]
  $$ e_1 \, \ffun{icf} \, e_2 \ \triangleq\ \exists e, \, e = \fpred(e_1) = \fpred(e_2)  $$
\end{definition}
Next, one can define the conflict relation as a consistent closure of $\ffun{icf}$
\begin{definition}[conflict relation]
  $$ e_1 \, \# \, e_2 \ \triangleq\ \exists e_3\, e_4, e_3 \, \ffun{icf} \, e_4 \ and \ e_3 \leqslant e_1, \, e_4 \leqslant e_2 $$
\end{definition}
However, there is one problem with such definition. It is clear that the relation defined in this way turns out to be symmetric and consistent. Will it be irreflexive? It turns out that in the general case, no. Consider the following program:
\input{fig/encode_prog.tex}

Let's build, as in the Step-by-step example, a fragment of $exec\_event\_structure$ corresponding to this program before reading from $y$
\input{fig/encode_part.tex}

One can observe, that our program in the first thread was divided into two independent executions: execution, in which we read from $x$ value 1 (denote by execution (a)) and execution, in which we read from $x$ 2 (execution (b)).

Now we want to add a read-from-$y$ event to our event structure. Let's see what is going to happen if we will add "read"-event, reading from $\wlab{y}{2}$ (see \cref{fig:cf}).
\input{fig/encode_cf.tex}

@@

To deal with this program, we need to impose a constraint on the event structure.

We will say that the event structure is consistent if there are no reads from conflicting writes. Formally:
\begin{definition}[consistent Execution Event Structure]
  $$consitent\langle E, \ffun{lab}, \fpred, \frf \rangle \ \triangleq \ \forall e \in E, \ \neg (e \ \# \ \frf e)$$
\end{definition}
It turns out that this is enough for the conflict relation to be irreflexive. (see \cref{fig:full})
\begin{lemma}
  Let $E$ be a consistent Execution Event Structure (denote as $cexec\_event\_structure$). Then it's conflict relation $\#$ is irreflexive
\end{lemma}
\input{fig/encode_full.tex} 
So, we see that from the consistent event structure one can get the Prime Event Structure

### 2021-03-24 papers/SYRCoSE/regmachine.tex
SYRCoSE: add some text

\subsection{Register Machine and thread semantics}
Now we are going to define language of simple register machine with shared memory.
\[
\begin{array}{rcll}
  \texttt{P} &::=& \texttt{ST}^* \\
  \texttt{ST} & ::= & \texttt{I}^* \\
  \texttt{I} & ::= & \texttt{Reg := Val} \\
                 & | & \texttt{Loc := Val} \\
                 & | & \texttt{Reg := Loc} \\
                 & | & \texttt{CJmp Reg $\mathbb{N} \backslash \{0\}$} \\
  \texttt{Reg}            & ::= & r \, | \, r' \, | ... \\ 
  \texttt{Loc}            & ::= & x \, | \,y \,| ... \\ 
\end{array}
\] where
\begin{itemize}
  \item \texttt{P} -- program is a sequence of single thread progams
  \item \texttt{ST} -- single thread program is a sequence of instructions
  \item \texttt{I} -- instruction can be either write to some register of some value or location value, either write to some location or conditional jump
\end{itemize}
Here \texttt{Reg} are registers (denoted as $r,r',...$) ~--- variables that should be different for different single thread progams. In contrast \texttt{Loc} are local variables (denoted as $x,y,...$) ~--- this is our shared (between different threads) memory. Also \texttt{Val} is the type of values (denoted as $i,j,...$) in our program, for simplicity let it be $\mathbb{N}$ or $\mathbb{Z}$. \\
We are interested at looking after shared memory, so in our event structures we have events that represent actions only at local variables. \\
Next we define semantics of single thread program. Let's consider some single-thread program (\texttt{ST} in our notation) and define thread state:
\begin{definition}[Thread State]
  $$thread\_state \ \triangleq \ \langle \ffun{ip}, \regmap \rangle,$$ where
  \begin{itemize}
    \item $\ffun{ip}$ ~--- our place in program (ie. line number) 
    \item $\regmap$ ~--- map from registers to values
  \end{itemize}
\end{definition}
If we have some $p : \texttt{ST}$, and some line number $\ffun{ip}$ we will denote by $p[\ffun{ip}]$ -- $\ffun{ip}$'s instuction of $p$ (we start numeration from 1).  \\
Then we define one-step semantics relation $st, p \xRightarrow{l} cont\_st$, where
\begin{itemize}
  \item $st$ ~--- thread state
  \item $p$ ~--- our single-thead programm
  \item $l$ ~--- label that denotes action on shared memory (can be with hole)
  \item $cont\_st$ ~--- continuation of thread state (ie. function from \texttt{Val} to $thread\_state$). We clarify why we need continuation here a bit later
\end{itemize}
If we have some thread state $st$, we can identify $\lambda\, x,\, st$ (ie. constant continuation) with $st$ itself. Also by $f[r \leftarrow i]$ we will denote function that sends $r$ to $i$ and $r'$ to $f(r')$, for all $r' \neq r$

@@

Every rule in \cref{fig:thread_sem} should be clear (see [TODO]), except the last one. Consider we want to read some local variable $x$ in some thread. Another threads can write something to this $x$ too, and it is possible to read from some of these writes. While building semantics of our whole programm we will have enough information to decide from wich write we can read. But for now let us consider that our side-effect is the label with the hole, and our result will be a continuation of a thread state (ie. some function $cont\_st : \texttt{Val} \to thread\_state$).

>>>>

\subsection{Register Machine and thread semantics}
Now we are going to define language of simple register machine with shared memory.
\[
\begin{array}{rcll}
  \texttt{P} &::=& \texttt{ST}^* \\
  \texttt{ST} & ::= & \texttt{I}^* \\
  \texttt{I} & ::= & \texttt{Reg := Val} \\
                 & | & \texttt{Loc := Val} \\
                 & | & \texttt{Reg := Loc} \\
                 & | & \texttt{CJmp Reg $\mathbb{N} \backslash \{0\}$} \\
  \texttt{Reg}            & ::= & r \, | \, r' \, | ... \\ 
  \texttt{Loc}            & ::= & x \, | \,y \,| ... \\ 
\end{array}
\] where
\begin{itemize}
  \item \texttt{P} -- program is a sequence of single thread progams
  \item \texttt{ST} -- single thread program is a sequence of instructions
  \item \texttt{I} -- instruction can be 
  \begin{itemize}
    \item \texttt{Reg := Val} -- writing some value to the register
    \item \texttt{Reg := Loc} -- reading a value from the shared-memory-location and writing it to the register
    \item \texttt{Loc := Val} -- writing some value to the shared-memory-location
    \item CJmp Reg $\mathbb{N} \backslash \{0\}$ -- conditional jump
  \end{itemize}
\end{itemize}
Here \texttt{Reg} are registers (denoted as $r,r',...$) ~--- variables that should be different for different single thread progams. In contrast \texttt{Loc} are local variables (denoted as $x,y,...$) ~--- this is our shared (between different threads) memory. Also \texttt{Val} is the type of values (denoted as $i,j,...$) in our program, for simplicity let it be $\mathbb{N}$ or $\mathbb{Z}$. \\
We are interested at looking after shared memory, so in our event structures we have events that represent actions only at local variables. \\
Next we define the semantics of a single thread program. Let's consider some single-thread program (\texttt{ST} in our notation) and define a thread state:
\begin{definition}[Thread State]
  $$thread\_state \ \triangleq \ \langle \ip, \regmap \rangle,$$ where
  \begin{itemize}
    \item $\ip$ ~--- our place in program (ie. line number) 
    \item $\regmap$ ~--- map from registers to values
  \end{itemize}
\end{definition}
If we have some $p : \texttt{ST}$, and some line number $\ip$ we will denote by $p[\ip]$ -- $\ip$'s instuction of $p$ (we start numeration from 1).  \\
Then we define one-step semantics relation $\st, p \xRightarrow{l} \cst$, where
\begin{itemize}
  \item $\st$ ~--- thread state
  \item $p$ ~--- our single-thead programm
  \item $l$ ~--- label that denotes action on shared memory (can be with hole)
  \item $\cst$ ~--- continuation of thread state (ie. function from \texttt{Val} to $thread\_state$). We clarify why we need continuation here a bit later
\end{itemize}
If we have some thread state $\st$, we can identify $\lambda\, x,\, \st$ (ie. constant continuation) with $\st$ itself. Also by $f[r \leftarrow i]$ we will denote function that sends $r$ to $i$ and $r'$ to $f(r')$, for all $r' \neq r$

@@

Every rule in \cref{fig:thread_sem} should be clear (see [TODO]), except maybe the last one. Consider we want to read some local variable $x$ in some thread. Another threads can write something to this $x$ too, and it is possible to read from some of these writes. While building semantics of our whole programm we will have enough information to decide from wich write we can read. But for now let us assume that our side-effect is the label with the hole, and our result is a continuation of a thread state (ie. some function $\cst : \texttt{Val} \to thread\_state$).

### 2021-03-24 papers/SYRCoSE/small_step.tex
SYRCoSE: add some text

\subsection{Construction small-step semantics}
Consider we have some Execution Event Structure $es = \la E, \ffun{lab}, \fpred, \frf \ra$, $pr \in E$ ~--- the $es$'s event and some fresh identifier $e = fresh(E)$. Let's define expansion $es$ with the event $e$ depending on its label:
\begin{definition}[Add event]
  \begin{itemize}
    \item $es + \la \wlab{x}{i}, \ pr \ra \ \triangleq \vspace{1mm} \\ \ 
    \la E \cup \{e\}, \
        \ffun{lab}[e \leftarrow \wlab{x}{i}], \
        \fpred[e \leftarrow pr], \
        \frf \ra\ $
        \vspace{1mm}
    \item $es + \la \text{ThreadStart} \ra \ \triangleq \vspace{1mm} \\
    \la E \cup \{e\}, \
    \ffun{lab}[e \leftarrow \text{ThreadStart}], \
    \fpred, \
    \frf \ra\ $
    \vspace{1mm}
    \item $es + \la \rlab{x}{i}, \ pr, \ wr \ra \ \triangleq$
    $$\la E \cup \{e\}, \
    \ffun{lab}[e \leftarrow \rlab{x}{i}], \
    \fpred[e \leftarrow pr], \
    \frf[e \leftarrow wr] \ra\ $$
  \end{itemize}
\end{definition}
Then we define programm configuration:
\begin{definition}[configuration]
  $$config \ \triangleq \ \la \es, \tmap \ra$$
  where
  \begin{itemize}
    \item $\es$ ~--- our current consistent Execution Event Structure
    \item $\tmap$ ~--- mapping from events in $\es$ to the pair of corresponding [TODO]
  \end{itemize}
\end{definition}

@@

\input{fig/small_step.tex}
[TODO: small step rules explanation]

>>>>

\subsection{Construction small-step semantics}
Consider we have some Execution Event Structure $es = \la E, \ffun{lab}, \fpred, \frf \ra$, $pr \in E$ ~--- the $es$'s event and some fresh identifier $e = fresh(E)$. Let's define expansion $es$ with $e$ depending on it's label:
\begin{definition}[Add event]
  \begin{itemize}
    \item $es + \la \wlab{x}{i}, \ pr \ra \ \triangleq \vspace{1mm} \\ \ 
    \la E \cup \{e\}, \
        \ffun{lab}[e \leftarrow \wlab{x}{i}], \
        \fpred[e \leftarrow pr], \
        \frf \ra\ $
        \vspace{1mm}
    \item $es + \la \text{ThreadStart} \ra \ \triangleq \vspace{1mm} \\
    \la E \cup \{e\}, \
    \ffun{lab}[e \leftarrow \text{ThreadStart}], \
    \fpred, \
    \frf \ra\ $
    \vspace{1mm}
    \item $es + \la \rlab{x}{i}, \ pr, \ wr \ra \ \triangleq$
    $$\la E \cup \{e\}, \
    \ffun{lab}[e \leftarrow \rlab{x}{i}], \
    \fpred[e \leftarrow pr], \
    \frf[e \leftarrow wr] \ra\ $$
  \end{itemize}
\input{fig/small_step.tex}
\end{definition}
Then we define programm configuration:
\begin{definition}[configuration]
  $$config \ \triangleq \ \la \es, \tmap \ra$$
  where
  \begin{itemize}
    \item $\es$ ~--- our current consistent Execution Event Structure
    \item $\tmap$ ~--- map that takes event $e$ from $\es$ and returns some pair. First component of this pair is a state of some thread after execution of $e$'s memory-side-effect. Our programm contains several threads, so we need to specify number of thread (where everything is going on) in the second component
  \end{itemize}
\end{definition}
So programm configuration describes current Execution Event Structure, where we map each event to corresponding thread state. Thus, to make our semantics step we should select some event and make progress in it's thread state. Before putting it formally we will need some notations:

@@

One can find definition of small-step semantics in \cref{fig:smallstep} \\
First look at $\to$ThreadStart rule. In this rule we are starting a new thread by adding some fresh identifier to $\es.E$ with label -- ThreadStart. \\
If we proceeded in thread state of some event $e$ without any memory-side-effect, we should just update $e$'s thread state (see $\to\_$). \\
Another possibility of memory-side-effect is write to some shared-memory location. In this case we should just add a new write-event to our Execution Event Structure, and map it to the thread state after our side-effect (see $\to\wlab{x}{i}$). \\
We observe the most interesting situation when our side-effect was $\rlab{x}{\_}$ (see $\to\rlab{x}{i}$). That means that we wanted to read from some shared-memory location. We need to choose some write-event $w \in \es.E$ and read value from it. It's clear that $w$ should have label $\wlab{x}{i}$, for some $i$. \\
Unfortunately, in the last step we could read from conflicting event $w$. So after making $\to$ step, it is necessary to chek if our final $\es'$ is consistent (see small-step rule). After all, we obtain one-step relation $\Rightarrow$.
